---
title: "B210533_assessment_R_markdown_file"
author: "B210533"
date: "19/06/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Working with data types and structures in Python and R
## Introduction
This is a project created as for the Working with data types and structures in Python and R (2021-2022)[FLEX] course at University of Edinburgh. Prepared and submitted in June 2022. 

##### Github repository
The data, code and analysis for this project can be accessed at: 
https://github.com/Exam-B210533/B210533_assessment 



# Loading NHSRdatasets
The data used in this project is from the NHSRdatasets package. This package has been created to support skills development in the NHS-R community and contains several free datasets. The dataset I have chosen is from the NHSRdatasets package and is a provisional record of deaths registered weekly in England and Wales from January 2010 to end of March 2020. It is contained in the ons_mortality dataset. 

All of the variables in the dataset were needed for my data capture tool, including categorisations (category 1 and category 2), counts, date and week number, and the data was subsetted into test and training data. The Jupyter Notebook "./ipynbScripts/CollectingDataUsingInteractiveJupyterWidgets.ipynb" was used to to collect the data.

### Load packages and data
Load the packages and data needed for this script.
```{r a,message = FALSE, warning = FALSE}
library(NHSRdatasets)
library(tidyverse)
library(here)
library(knitr)
library(lubridate)
library(caret)
```
## ONS Mortality Data 
This project uses the ons_mortality data from the NHSRdatasets package.

```{r b,message = FALSE, warning = FALSE}
data(ons_mortality)
mort <- ons_mortality
class(mort)
glimpse(mort)
```
This is the output of the `glimpse()’ function. It starts off with the number of rows and columns and each column in separate rows. The Category 1 and Category 2 columns have multiple different character values and these should be examined to see the unique values in the set. 

### Unique Categories

The unique values for these are are follows: 

**Category 1 unique values**
```{r c,message = FALSE, warning = FALSE}
unique(mort[c("category_1")])
```
**Category 2 unique values**
```{r d,message = FALSE, warning = FALSE}
unique_cat2 = unique(mort[c("category_2")])
print(unique_cat2, n=42)
```
Using these unique values its possible to see that the dataset contains:

  * **category_1:** character, containing the names of the groups for counts, e.g. "Total deaths", "all ages". 
    + Total deaths                                                                         
    + All respiratory diseases (ICD-10 J00-J99) ICD-10                                     
    + Persons                                                                              
    + Males                                                                                
    + Females                                                                              
    + Region                                                                               
    + NA                                                                                   
    + average of same week over 5 years                                                    
    + Deaths where COVID-19 was mentioned on the death certificate (ICD-10 U07.1 and U07.2)
  * **category_2**: character, subcategory of names of groups where necessary, e.g. details of region: "East", details of age bands "15-44". Can be: 
    + all ages                         
    + average of same week over 5 years
    + v2001       
    + v2010
    + v2013
    + NA
    + Under 1 year                     
    + 01-14                            
    + 15-44                            
    + 45-64                            
    + 65-74                            
    + 75-84                            
    + 85+ 
    + North East
    + North West
    + Yorkshire
    + East Midlands
    + West Midlands
    + East
    + London
    + South East 
    + South West
    + Wales
    + 1-4
    + 5-9
    + 10-14
    + 15-19
    + 20-24
    + 25-29
    + 30-34
    + 35-39
    + 40-44
    + 45-49
    + 50-54
    + 55-59
    + 60-64
    + 65-69
    + 70-74
    + 75-79
    + 80-84
    + 85-89
    + 90+
  * **counts:** numeric, numbers of deaths in whole numbers and average numbers with decimal points. To retain the integrity of the format this column data is left as character.
  * **date:** date, format is yyyy-mm-dd; all dates are a Friday.
  * **week_no:**  integer, each week in a year is numbered sequentially.
  * **performance:** the performance ([1 - breaches]/attendances) calculated for the whole of England.
  * **consent:** the consent from the end-user to process and share the data collected with the data capture tool.


### Range of Dates

The dataset contains values for dates between: 
```{r e,message = FALSE, warning = FALSE}
min(mort$date, na.rm = TRUE)
```
minimum and
```{r f, message = FALSE, warning = FALSE}
max(mort$date, na.rm = TRUE)
``` 
maximum

### Missing Values

The list of unique values for Category 1 and Category 2 shows that they contain NA values. It's also necessary to examine if other columns contain missing values. 

```{r g, message = FALSE, warning = FALSE}
mort %>% 
  map(is.na) %>%
  map(sum)
```
**counts** also contains missing values. It will be necessary to recheck this once the subset of data has been created to see if these missing values are also present in the subset of data. 

### Add an index to the data set

It's necessary to separate the data into training and testing dta sets. The test data will be used to evaluate data collection and analysis tools. A data index column is added to the raw data so that partitioned data sets can be linked back to the raw data if needed in the future.

```{r h, message = FALSE, warning = FALSE}
mort <- rowid_to_column(mort, "index")

```

### Tabulute the raw data

The raw data with index is then tabulated to make the data more readable. 

```{r i, message = FALSE, warning = FALSE}
mort %>%
  # Set the period column to show in month-year format
  mutate_at(vars(date), format, "%b-%y") %>% 
  # Set the numeric columns to have a comma at the 1000's place
  mutate_at(vars(counts), comma) %>%
  # Show the first 10 rows
  head(10) %>%
  # Format as a table
  kable()

```


### Save the raw data to the project

The raw data is saved to the project folder using the here()) function. 

```{r j, message = FALSE, warning = FALSE}
write_csv(mort, here("RawData", "ons_mortality.csv"))

```

# Selecting variable for analysis

This project does not require all of the data contained in ons__mortality. A subset of the data will be created for use in the project. This project will examine the variance between the actual mortality and the 5-year average by season. This will be done for data on Total deaths. A subset of data is needed containing "Total Deaths" for "All Ages" and the "average of same week over 5 years". 

### Examine the lengths of datasets to be mapped

First, filter() is used to create an object with the "average of the same week over 5 years" data.

```{r k, message = FALSE, warning = FALSE}
mort_5year_avg <- filter(mort,category_2 == "average of same week over 5 years")
glimpse(mort_5year_avg)
tail(mort_5year_avg)
```

There are **521 rows** of data. The records exist only for Category 1 "Total Deaths" and run from Week 1 (08/01/2010) to Week 52 (27/12/2019). Next, the filter() function is used to create an object with only "Total deaths" and "all ages" to examine the length and dates of Total deaths and all ages data. 

```{r l, message = FALSE, warning = FALSE}
mort_total <- filter(mort,category_1 == "Total deaths" & category_2 == "all ages")
glimpse(mort_total)
```

The total deaths for all ages has **535 observations** while the average data has **521 observations**. The last date for 5-year average observation is 27-12-2019, so need to remove "all deaths" observations after this date. Now we create a subset of the data for "all deaths" and "all ages" up to and including the last week of December 2019.

```{r m, message = FALSE, warning = FALSE}
mort_all <- filter(mort,category_1 == "Total deaths" & category_2 == "all ages" & date < "2019-12-31")
glimpse(mort_all)
```

## Append the 5 year average

The 5 years average data for the relevant dates is appended as a column to the subset of Total deaths for all ages.

```{r n , message = FALSE, warning = FALSE}
mort_all$mort_avg = mort_5year_avg$counts
```

## Calculate the variance between actual deaths and the 5 year average

Another column is added with a new variable with the calculated difference between the actual mortality and the average for the 5 years preceding that date. 

```{r o, message = FALSE, warning = FALSE}
mort_all$variance_from_avg = mort_all$counts / mort_all$mort_avg
```

## Create a year column

The data will be examined and plotted using the week and year for the observation. For this purpose it is helpful to have a year variable based on the date for the same observation. 

```{r p, message = FALSE, warning = FALSE}
mort_all$year = year(mort_all$date)
```

# Create the subset data frame

The final data frame for analysis is created. It includes the data objects:
  * index 
  * week number
  * year 
  * date
  * count of deaths
  * average of previous 5 years
  * variance between actual count and average. 

```{r q, message = FALSE, warning = FALSE}
mort_all <- mort_all[, c("index", "week_no", "year", "date", "counts", "mort_avg", "variance_from_avg")]
head(mort_all)
```

## Check again for missing values

There were NA values in teh count column when the raw data was examined. Now its necessary to check again to see if the subset is complete or whether null values need to be managed. 

```{r r, message = FALSE, warning = FALSE}
mort_all %>% 
  map(is.na) %>%
  map(sum)
```

There are no missing values and so we can proceed with analysis. 

# Tabulate the subset 

The subset data is then tabulated to make the data more readable. 

```{r s, message = FALSE, warning = FALSE}
mort_all %>%
  # set the numeric columns to have a comma at the 1000's place
  mutate_at(vars(counts, mort_avg), comma) %>%
  # reduce decimal places 
  mutate_at(vars(variance_from_avg), round, 2)  %>%
  # show the first 10 rows
  head(10) %>%
  # format as a table
  kable(caption= "England Mortality Data and Variance from Average, 2010 - 2019")

```
# Save the subset to the project
The subset of data is saved to the project folder using the here()) function. 

```{r t, message = FALSE, warning = FALSE}
write_csv(mort_all, here("RawData", "ons_mortality_ENG_1019.csv"))
```
# Partitian the data 
A subset of the data is needed for use in evaluation of the data capture tool. 

### Calculate the proportion of the data subset to use for testing purposes
```{r u, message = FALSE, warning = FALSE}
prop<-(1-(15/nrow(mort_all)))
print(prop)
```
### Set a seed to be able to replicate the random generator in future
The 'set.seed()' function is a random number generator, which is useful for creating random objects that can be reproduced. This will make sure that every time we run this script, we will partition the raw data into the same test and training data.

```{r v, message = FALSE, warning = FALSE}
set.seed(432)
```
### Partition the raw data 
The createDataPartition function is used to generate an object using a proportion of index rows. The index object is used to create the training set. 

```{r w, message = FALSE, warning = FALSE}
train_index <- createDataPartition(mort_all$index, p = prop, 
                                  list = FALSE, 
                                  times = 1)
head(train_index)
```
### Assign the training data to a new object

All records in the train_index are assigned to the training data.

```{r x, message = FALSE, warning = FALSE}
mort_alltrain <- mort_all[ train_index,]
nrow(mort_alltrain)
```
There are 509 records in the training set

### Tabulate the training dataset

```{r y, message = FALSE, warning = FALSE}
mort_alltrain %>%
  # set the numeric columns to have a comma at the 1000's place
  mutate_at(vars(counts, mort_avg), comma) %>%
  # reduce decimal places 
  mutate_at(vars(variance_from_avg), round, 2)  %>%
  # show the first 10 rows
  head(10) %>%
  # format as a table
  kable(caption= "Training Data: England Mortality Data and Variance from Average, 2010 - 2019")
```

### Save the raw data to the project

The raw data is saved to the project folder using the here()) function. 

```{r z, message = FALSE, warning = FALSE}
write_csv(mort_alltrain, here("Data", "ons_mortality_ENG_1019_training.csv"))
```

### Create the test subset

Allocate all records that are not assigned to the training data subset

```{r aa, message = FALSE, warning = FALSE}
mort_alltest  <- mort_all[-train_index,]
nrow(mort_alltest)
```
There are 12 records in the test data. One row needs to be set aside for use by the assignment markers to test and evaluate the data-capture tool.

### Create the Markers single-row test subset

```{r ab, message = FALSE, warning = FALSE}
mort_alltestMarker  <- mort_alltest[1,]
```

### Tabulate the Markers row for testing

The subset data is then tabulated to make the data more readable. 

```{r ac, message = FALSE, warning = FALSE}
mort_alltestMarker %>%
  # set the numeric columns to have a comma at the 1000's place
  mutate_at(vars(counts, mort_avg), comma) %>%
  # reduce decimal places 
  mutate_at(vars(variance_from_avg), round, 2)  %>%
  # show the first 10 rows
  head(10) %>%
  # format as a table
  kable(caption= "Markers Test Data: England Mortality Data and Variance from Average, 2010 - 2019")
```
### Save the raw data to the project

The raw data is saved to the project folder using the here()) function. 

```{r ad, message = FALSE, warning = FALSE}
write_csv(mort_alltestMarker, here("Data", "ons_mortality_ENG_1019_test_marker.csv"))
```

